---
title: "05_analysis_1"
author:
  - "Kacper Maciejewski (s243548)"
  - "Sofia Russo (252054)"
  - "Aleksandra Wozniak (253713)"
  - "Gabriel Loayza (252608)"
  - "Dagmar Geevers (252256)"
format:
  html:
    embed-resources: true
editor: visual
---

## Load the data

```{r}

library(tidyverse)
library(pheatmap)
library(tidyr)
library(broom)
library(grid)

my_data <- read_csv("../data/02_dat_clean.csv")
view(my_data)
```

## Checking the data

Grade is a microscopic assessment of the appearance of cancer cells, indicating how different they are from the normal cells.

-   low grade- the cells are similar to healthy ones. The tumor is usually less agressive

-   high grade- the cells are poorly differentiated or undifferentiated. The tumor is usually more agressive

```{r}
print(unique(my_data$Grade))
#now we are replacing text with numbers
changed_data <- my_data %>% 
  mutate(
    Grade_in_numbers = case_when(
#group 1: low malignacy
      Grade == "G1" ~ 1,
      Grade == "Low Grade" ~ 1,
#group 2: medium malignacy
      Grade == "G2" ~ 2,
#group 3: high malignacy
      Grade == "G3" ~ 3,
      Grade == "High Grade" ~ 3,
#group 4: the highest malignacy
      Grade == "G4" ~ 4,
      TRUE ~ NA_real_
    )
  ) %>% 
  drop_na(Grade_in_numbers)
print(changed_data$Grade_in_numbers)
```

## Correlation

We want to see how the activity of gene x in patients with cancer y correlates with increasing malignancy

For that we are using Spearman correlation, because it is resistant for asymmetry and it looks at ranks

```{r}

changed_data$Cancer <- dplyr::recode(changed_data$Cancer,
  "blca_tcga_pan_can_atlas_2018"    = "BLCA",
  "brca_tcga_pan_can_atlas_2018"    = "BRCA",
  "cesc_tcga_pan_can_atlas_2018"    = "CESC",
  "coadread_tcga_pan_can_atlas_2018"= "COAD",
  "esca_tcga_pan_can_atlas_2018"    = "ESCA",
  "gbm_tcga_pan_can_atlas_2018"     = "GBM",
  "hnsc_tcga_pan_can_atlas_2018"    = "HNSC",
  "kirc_tcga_pan_can_atlas_2018"    = "KIRC",
  "kirp_tcga_pan_can_atlas_2018"    = "KIRP",
  "lgg_tcga_pan_can_atlas_2018"     = "LGG",
  "lihc_tcga_pan_can_atlas_2018"    = "LIHC",
  "luad_tcga_pan_can_atlas_2018"    = "LUAD",
  "lusc_tcga_pan_can_atlas_2018"    = "LUSC",
  "ov_tcga_pan_can_atlas_2018"      = "OV",
  "paad_tcga_pan_can_atlas_2018"    = "PAAD",
  "prad_tcga_pan_can_atlas_2018"    = "PRAD",
  "sarc_tcga_pan_can_atlas_2018"    = "SARC",
  "skcm_tcga_pan_can_atlas_2018"    = "SKCM",
  "stad_tcga_pan_can_atlas_2018"    = "STAD",
  "tgct_tcga_pan_can_atlas_2018"    = "TGCT",
  "thca_tcga_pan_can_atlas_2018"    = "THCA",
  "ucec_tcga_pan_can_atlas_2018"    = "UCEC"
)

  
cancer_order <- c(
  "LIHC", "UCEC", "KIRC", "LGG",
  "OV", "PAAD",
  "HNSC", "ESCA", "CESC", "STAD",
  "COAD", "BLCA"
)
cancers_use <- intersect(cancer_order, unique(changed_data$Cancer))

changed_data <- changed_data %>%
  filter(Cancer %in% cancers_use)

changed_data$Cancer <- factor(changed_data$Cancer,
                              levels = cancers_use)

#here i define genes in the order for the graph
genes_to_test <- c(
#class I
  "HDAC3", "HDAC2", "HDAC1", "HDAC8",
#class IIA
  "HDAC7", "HDAC9", "HDAC5", "HDAC4",
#class III
  "SIRT1", "SIRT7", "SIRT6", "SIRT4", "SIRT3", "SIRT2", "SIRT5",
#class IIB
  "HDAC6", "HDAC10",
#class IV
  "HDAC11")


#correlation for each pair
#pivot_longer- from wide to long
s_correlation <- changed_data %>% 
  pivot_longer(
    cols = any_of(genes_to_test),
    names_to = "gene",
    values_to = "expression"
  ) %>% 
  #group for each pair (gene + cancer)
group_by(Cancer, gene) %>% 
summarize(
    summary_test = broom::tidy(cor.test(expression, Grade_in_numbers, method ="spearman", exact = FALSE)), .groups = "drop"
  ) %>% 
  unnest(summary_test) %>% 
  select(Cancer, gene, rho = estimate, p.value)

```

## Matrix_rho

heatmap requires data in matrix format

```{r}
matrix_rho <- s_correlation %>% 
  select(gene, Cancer, rho) %>% 
  pivot_wider(
    names_from = Cancer,
    values_from = rho
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix()
```

## Matrix_p

```{r}
matrix_p <- s_correlation %>% 
  select(gene, Cancer, p.value) %>% 
  pivot_wider(
    names_from = Cancer,
    values_from = p.value
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix()

#filtering and sorting- without that R will sort the genes by alphabet 
final_genes <- genes_to_test[genes_to_test %in% rownames(matrix_rho)]
final_cancers <- cancers_use[cancers_use %in% colnames(matrix_rho)]

#sorted versions
matrix_rho <- matrix_rho[final_genes, final_cancers, drop = FALSE]
matrix_p   <- matrix_p[final_genes, final_cancers, drop = FALSE]

matrix_rho_plot <- matrix_rho
matrix_rho_plot[is.na(matrix_rho_plot)] <- 0
```

## Pheatmap

From the correlation:

-   Positive numbers - the more malignant the cancer, the more active the gene (red)

-   Negative numbers - when the cancer is more malignant, the gene is less active (blue)

-   Numbers close to 0 - no correlation

-   Matrix_p says how random are the results. On a graph a dot means that the value is small, so the result is certain

```{r}
#adnotations-> annot_row creates a legend, assignment to classes
gene_classes <- c(
  HDAC3="class I",
  HDAC2="class I",
  HDAC1="class I",
  HDAC8="class I",

  HDAC7="class IIA",
  HDAC9="class IIA",
  HDAC5="class IIA",
  HDAC4="class IIA",
  
  SIRT1="class III",
  SIRT7="class III",
  SIRT6="class III",
  SIRT4="class III",
  SIRT3="class III",
  SIRT2="class III",
  SIRT5="class III",
  
  HDAC6="class IIB",
  HDAC10="class IIB",

  HDAC11="class IV"
)
#data frame for heat map
annot_row <- data.frame(
  HDAC_family = as.character(gene_classes[final_genes]),
  row.names = final_genes,
  stringsAsFactors = FALSE)

annot_row$HDAC_family <- factor(
  annot_row$HDAC_family,
  levels = c("class I", "class IIA", "class IIB", "class III", "class IV")
)
#defining colors for each class: red, blue, green, purple, orange
ann_colors = list(
  HDAC_family = c(
    "class I" = "salmon",   
    "class IIA" = "#E69F52", 
    "class IIB" = "#BDB600", 
    "class III" = "#009E60", 
    "class IV" = "#008ECE"))
    
#BH correction for many tests- reduces fake positives 
matrix_p_dot <- matrix(p.adjust(as.vector(matrix_p), method = "BH"),
                       ncol = ncol(matrix_p),
                       nrow = nrow(matrix_p),
                       byrow = FALSE
)


display_matrix <- 
  ifelse(matrix_p_dot < 0.001, "■",
  ifelse(matrix_p_dot < 0.01,  "▲",
  ifelse(matrix_p_dot < 0.05,  "•", "")))

my_breaks <- seq(-0.3, 0.3, length.out = 101)

my_colours <- colorRampPalette(c("blue","white", "red"))(100)


hm <- pheatmap(
  matrix_rho_for_plot,
  color = my_colours,
  breaks = my_breaks,
  na_col = "grey90",
  cluster_rows = FALSE,
  cluster_cols = FALSE, #cancers in teh same order as in article on a graph
  annotation_row = annot_row, 
  annotation_colors = ann_colors,
  display_numbers = display_matrix,
  fontsize_number = 6,
  main = " ",
  number_color = "black",
  show_colnames = TRUE,
  show_rownames = TRUE,
  annotation_legend = TRUE,
  fontsize = 5,
  fontsize_col = 8,
  border_color = NA,
  cellwidth = 13,
  cellheight = 10
  )



```

```         
```
